Does a fully convective star need a special corona to produce a polar flare?

### 2023-08-08 src/tex/ms.tex
[showyourwork] overleaf sync

\newcommand{\Tcool}{\input{output/T1.tex}}
\newcommand{\Thot}{\input{output/T2.tex}}

@@

\author[]{Beate Stelzer}
\affiliation{Institut für Astronomie & Astrophysik, Eberhard-Karls-Universität Tübingen, Sand 1, 72076 Tübingen, Germany}
\affiliation{INAF – Osservatorio Astronomico di Palermo, Piazza del Parlamento 1, 90134 Palermo, Italy}

@@

    We studied five Sectors of TESS observations, and obtained 36 ks ob XMM-Newton observations, to find that TIC 277's corona does not differ significantly from other stars with its spectral type and fast rotation period ($P_{\rm rot})$ with respect to X-ray luminosity, coronal temperatures, and flaring activity. Additionally, XMM-Newton captured a simultaneous X-ray and optical flare on TIC 277 with a bolometric energy of XXXX and a soft X-ray energy of XXXX in the 0.2-2 keV band. Combining the 18 flares found in TESS with the flare from XMM-Newton, we inferred that TIC 277 produces flares at or above this energy approximately once every 10 h. We suggest that TIC 277's X-ray and flaring activity is representative of other rapidly rotating late M dwarfs, but that further measurements of localized flares are needed to understand the conditions under which high latitude flares in these stars can occur.

@@

    \caption{TIC 277539431 (WISEA J105515.71-735611.3)}
    \begin{tabular}{ll}\hline 
         Parameter & Value  \\\hline
         Distance $d$ & $13.70\pm0.11\,$pc [2] \\
         % $G$ mag & $14.7189 \pm 0.0011$ [4]\\
         % $G_{BP}$ mag & \\
         % $G_{BP}$ mag & \\
         % $K_s$ mag & $9.666 \pm 0.024$ [3]\\
         % $J$ mag & $10.630 \pm 0.023$ [3] \\
         SpT & M7 [1]\\
         $T_{eff}$ & $2680^{60}_{-50}\,$ K [5]\\
         % age & \\
         Rotation period $P$ & $273.618 \pm 0.007\,$min [1]\\
         $v\sin i$ & $38.6\pm1.0\,$km/s [1] \\
         Inclination $i$ & $87.0^{+2.0}_{-2.4}\,$ deg [1]\\
         Radius $R$ & $0.145\pm0.004\,R_\odot$ [1]\\\hline

@@

From the XMM-Newton EPIC spectra, we could measure the coronal properties of TIC in both its quiescent, and flaring states~(Table~\ref{tab:specfit}). Using the combined quiescent and flaring data, we found a two-temperature VAPEC model to fit better than a one-temperature solution and no worse than a three-temperature solution. The resulting fit~(Fig.~\ref{fig:spec_joint_all}) consists of a cooler and a hotter component of about \Tcool and \Thot, respectively. With these temperatures as ground truth, we also fitted two-temperature VAPEC models to the flaring and quiescent portions of the observations separately. The coronal temperatures did not update significantly using either subset, although the quiescent spectrum appeared slightly cooler overall. However, the emission measure weighted mean temperature was about $1.5\,$MK cooler in the quiescent spectrum, and $\sim3.2\,$MK hotter in the flaring spectrum. This suggests that the flaring corona emits the hotter component, while the cooler component corresponds to the quiescent corona. In this interpretation, the presence of a hot component in the quiescent spectrum indicates that TIC 277 is flaring below the detection threshold, consistent with an extrapolation of the flare frequency distribution to lower temperatures~(see the following Section, and Fig.~\ref{fig:ffd}).

@@

From the PN spectra, we measured a soft X-ray flux of $F_X=XXXX$, and a luminosity $L_X=XXXX$ in the $0.2-2\,$keV band. 
 \begin{table}
% \movetableright=-20mm
\centering
    \script{TABLE_specfits.py}
    \caption{XSPEC fits to EPIC spectra for different subsets of observations.}
    \input{output/mcmc_specfit.tex}
        \label{tab:specfit}
\end{table}

@@

The FFD power law fit converged after 13,500 steps on a slope $\alpha = 1.9_{-0.2}^{+0.2}$, and $\log_{10}\beta=27.7_{-6.0}^{+7.1}\,\mathrm{d}^{-1}$. The frequency $R_{31.5}$ of flares per day above $\log_{10} E = 31.5\,\mathrm{erg}$ is about $-1.1$. The slope is typical of other flaring stars, regardless of spectral type a rotation period~(XXXX). $R_{31.5}$ is typical of flaring M dwarfs, in particular typical of late M dwarfs in the saturated activity regime, where $R_{31.5}$ becomes independent of rotation period~\citep{medina2020flare,murray2022study}.

For the X-ray flare in Fig.~\ref{fig:lightcurves}, we obtain a total energy of $6.5\pm0.7\cdot10^{30}\,$erg in the $0.2-5\,$keV band. The energy of the corresponding OM flare is $1.9\pm0.4\cdot 10^{30}\,$erg, which fits the extension of TESS' flare frequency distributions to lower energies within uncertainties on the power law fit~(Fig.~\ref{fig:ffd}). About one flare every $10\,$h of the order of the observed one or above can therefore be expected in future X-ray observations of TIC~277, assuming all optical flares above this energy in the OM data correspond to X-ray flares of similar or higher energy to the observed. \citet{guarcello2019simultaneous} find moderate correlation between flare energies in Kepler/K2 data and XMM-Newton, but the majority of their sampled energies are above $10^{33}\,$erg both in X-ray and optical. \citet{kuznetsov2021stellar} compare primary Kepler mission observations with XMM-Newton to find a similar result. We note that the X-ray flare light curve follows the typical fast-rise exponential decay shape, and is therefore unlikely eclipsed by the star~\citep{johnstone2012soft}.

@@

\subsection{X-ray luminosity}
\label{sec:discussion:xraylum}
Similar to its temperature make-up, while TIC~277's coronal luminosity of XXXX is somewhat lower than the $L_X/L_{\rm bol}\sim 10^{-3}$ average for saturated regime, it is not an outlier to other M dwarfs in this activity regime.

Using eROSITA data, \citet{magaudda2022first} cover both fully convective and rapidly rotating M dwarfs, and find a downward trend in $L_X/L_{\rm bol}$ with decreasing rotation period, pointing toward a supersaturation of the corona at fast rotation speeds, as has been conjectured before~\citep{jeffries2011investigating,ramsay2020tess}.   \citet{jeffries2011investigating} suggest that the cause for supersaturation is centrifugal stripping of the corona, but \citet{reiners2022magnetism} show that if magnetic field increases, X-ray XXXX increases likewise with no sign of a break. They conclude that the trend is rather due to the underlying dynamo. If TIC~277 is indeed underluminous, these results suggest that its dynamo produces a weaker magnetic field than that of other saturated M dwarfs, which results in a slightly lower X-ray luminosity. However, since both the temperature structure and flaring activity of TIC~277 are in line with saturated activity, this conjecture currently lacks evidence from other activity indicators. Moreover, late M dwarfs are underrepresented in these studies.

\citet{stelzer2022first} investigated late M and L dwarfs with eROSITA to show that $L_X/L_{\rm bol}$ increases from the full convection boundary (M3) towards later spectral types, but begins to decrease again from M7 downward. In that study, TIC~277 is at the lower boundary of the $L_X/L_{\rm bol}$ distribution for M7 dwarfs. At the same time, strong magnetic fields are found across the board in brown dwarfs~(XXXX). While it is not clear whether M7 falls in the stellar regime of~\citet{reiners2022magnetism} or into that latter brown dwarf realm, or has features of both, it appears at the lower bound regardless. The small number of supersaturated stars in \citet{magaudda2022first} tentatively suggests that this may be a common phenomenon among mid-to-late M dwarfs when they are quickly rotating.

>>>>

\newcommand{\Tcool}{\input{output/T1.tex}}
\newcommand{\Thot}{\input{output/T2.tex}}
\newcommand{\eom}{\input{output/om_flare.tex}}
\newcommand{\eepic}{\input{output/epic_flare.tex}}
\newcommand{\Tqmean}{\input{output/Tqmean.tex}}
\newcommand{\Tfmean}{\input{output/Tfmean.tex}}
\newcommand{\FX}{\input{output/epic_flux.tex}}
\newcommand{\LX}{\input{output/epic_Lx.tex}}
\newcommand{\Lbol}{\input{output/Lbol.tex}}
\newcommand{\ffdalpha}{\input{output/tess_ffd_alpha}}
\newcommand{\ffdbeta}{\input{output/tess_ffd_beta}}
\newcommand{\ffdr}{\input{output/R315}}

@@

\author{Beate Stelzer}
\affiliation{Institut für Astronomie \& Astrophysik, Eberhard-Karls-Universität Tübingen, Sand 1, 72076 Tübingen, Germany}
\affiliation{INAF – Osservatorio Astronomico di Palermo, Piazza del Parlamento 1, 90134 Palermo, Italy}

@@

    We studied five Sectors of TESS observations, and obtained 36 ks ob XMM-Newton observations, to find that TIC 277's corona does not differ significantly from other low mass stars on the saturated activity branch with respect to X-ray luminosity, coronal temperatures, and flaring activity. Additionally, XMM-Newton captured a simultaneous X-ray and optical flare on TIC 277 with a bolometric energy of \eom and a soft X-ray energy of \eepic in the 0.2-2 keV band. Combining the 18 flares found in TESS with the flare from XMM-Newton, we inferred that TIC 277 produces flares at or above this energy approximately once every 10 h. We suggest that TIC 277's X-ray and flaring activity is representative of other rapidly rotating late M dwarfs, but that further measurements of localized flares are needed to understand the conditions under which high latitude flares in these stars can occur.

@@

    \caption{TIC 277539431 (WISEA J105515.71-735611.3)}
    \begin{tabular}{ll}\hline 
         Parameter & Value  \\\hline
         Distance $d$ & $13.70\pm0.11\,$pc [2] \\
         % $G$ mag & $14.7189 \pm 0.0011$ [4]\\
         % $G_{BP}$ mag & \\
         % $G_{BP}$ mag & \\
         % $K_s$ mag & $9.666 \pm 0.024$ [3]\\
         % $J$ mag & $10.630 \pm 0.023$ [3] \\
         SpT & M7 [1]\\
         $T_{eff}$ & $2680^{60}_{-50}\,$ K [5]\\
         % age & \\
         Rotation period $P$ & $273.618 \pm 0.007\,$min [1]\\
         $v\sin i$ & $38.6\pm1.0\,$km/s [1] \\
         Inclination $i$ & $87.0^{+2.0}_{-2.4}\,$ deg [1]\\
         Radius $R$ & $0.145\pm0.004\,R_\odot$ [1]\\
         $L_{\rm bol}$ & \Lbol \\\hline

@@

From the XMM-Newton EPIC spectra, we could measure the coronal properties of TIC in both its quiescent, and flaring states~(Table~\ref{tab:specfit}). Using the combined quiescent and flaring data, we found a two-temperature VAPEC model to fit better than a one-temperature solution and no worse than a three-temperature solution. The resulting fit~(Fig.~\ref{fig:spec_joint_all}) consists of a cooler and a hotter component of about \Tcool and \Thot, respectively. With these temperatures as ground truth, we also fitted two-temperature VAPEC models to the flaring and quiescent portions of the observations separately. The coronal temperatures did not update significantly using either subset, although the quiescent spectrum appeared slightly cooler overall. However, the emission measure weighted mean temperature change dfrom \Tqmean in the quiescent spectrum to \Tfmean in the flaring spectrum. This suggests that the flaring corona emits the hotter component, while the cooler component corresponds to the quiescent corona. In this interpretation, the presence of a hot component in the quiescent spectrum indicates that TIC 277 is flaring below the detection threshold, consistent with an extrapolation of the flare frequency distribution to lower temperatures~(see the following Section, and Fig.~\ref{fig:ffd}).

@@

From the PN spectra, we measured a soft X-ray flux of \FX, and a luminosity \LX in the $0.2-2\,$keV band. 
 \begin{table}
% \movetableright=-20mm
\centering
    \script{TABLE_specfits.py}
    \caption{XSPEC fits to EPIC spectra for different subsets of observations.}
    \input{output/mcmc_specfit.tex}
        \label{tab:specfit}
\end{table}

@@

The FFD power law fit converged after 13,500 steps on a slope $\alpha =$\ffdalpha. The frequency $R_{31.5}$ of flares per day above $\log_{10} E = 31.5\,\mathrm{erg}$ is about $\log_{10}R_{31.5}=$\ffdr. The slope is typical of other flaring stars, regardless of spectral type a rotation period~(see, e.g., Fig.~13 in \citealt{ilin2021flares}). $R_{31.5}$ is typical of flaring M dwarfs, in particular typical of late M dwarfs in the saturated activity regime, where $R_{31.5}$ becomes independent of rotation period~\citep{medina2020flare,murray2022study}.

For the X-ray flare in Fig.~\ref{fig:lightcurves}, we obtain a total energy of \eepic in the $0.2-5\,$keV band. The energy of the corresponding OM flare is \eom, which fits the extension of TESS' flare frequency distributions to lower energies within uncertainties on the power law fit~(Fig.~\ref{fig:ffd}). About one flare every $10\,$h of the order of the observed one or above can therefore be expected in future X-ray observations of TIC~277, assuming all optical flares above this energy in the OM data correspond to X-ray flares of similar or higher energy to the observed. \citet{guarcello2019simultaneous} find moderate correlation between flare energies in Kepler/K2 data and XMM-Newton, but the majority of their sampled energies are above $10^{33}\,$erg both in X-ray and optical. \citet{kuznetsov2021stellar} compare primary Kepler mission observations with XMM-Newton to find a similar result. We note that the X-ray flare light curve follows the typical fast-rise exponential decay shape, and is therefore unlikely eclipsed by the star~\citep{johnstone2012soft}.

@@

\subsection{X-ray luminosity}
\label{sec:discussion:xraylum}
Similar to its temperature make-up, while TIC~277's coronal luminosity relative to bolometric luminosity of $L_X/L_{\rm bol}=\LXLbol$ is somewhat lower than the $L_X/L_{\rm bol}\sim 10^{-3}$ average in the saturated regime~\citep{wright2018stellar}, it is not an outlier to other M dwarfs in this activity regime.

In the Hyades and Praesepe, \citet{nunez2022factory} cover both fully convective and rapidly rotating M dwarfs, and find a downward trend in $L_X/L_{\rm bol}$ with decreasing rotation period for M3.5-M7 dwarfs, tentatively pointing toward a supersaturation of the corona at fast rotation speeds, as has been conjectured before~\citep{jeffries2011investigating,ramsay2020tess}.   \citet{jeffries2011investigating} suggest that the cause for supersaturation is centrifugal stripping of the corona, but \citet{reiners2022magnetism} show that if magnetic field increases, X-ray XXXX increases likewise with no sign of a break. They conclude that the trend is rather due to the underlying dynamo. If TIC~277 is indeed underluminous, these results suggest that its dynamo produces a weaker magnetic field than that of other saturated M dwarfs, which results in a slightly lower X-ray luminosity. However, since both the temperature structure and flaring activity of TIC~277 are in line with saturated activity, this conjecture currently lacks evidence from other activity indicators. Moreover, late M dwarfs are underrepresented in these studies.

\citet{stelzer2022first} investigated late M and L dwarfs with eROSITA to show that $L_X/L_{\rm bol}$ increases from the full convection boundary (M3) towards later spectral types, but begins to decrease again from M7 downward. In that study, TIC~277 is at the lower boundary of the $L_X/L_{\rm bol}$ distribution for M7 dwarfs. At the same time, strong magnetic fields are found across the board in brown dwarfs~(XXXX). While it is not clear whether M7 falls in the stellar regime of~\citet{reiners2022magnetism} or into that latter brown dwarf realm, or has features of both, it appears at the lower bound regardless. The small number of supersaturated stars in \citet{nunez2022factory}  can only tentatively suggest that this is a common phenomenon among mid-to-late M dwarfs when they are quickly rotating.
